Objectification in Action: Self- and Other-Objectification in Same-gender Interactions

### 2020-07-28 Final_Manuscript.tex
lab meeting changes

Roberts and Fredrickson (1997) Objectification Theory, suggests that
women are objectified in interpersonal encounters, and that the effect
might be strongest from men, but possibly also from women. Psychological
researchers studying the sexual objectification of women, and the
resulting self-objectification that follows, have in recent years
explored the interpersonal process of the objectification -
self-objectification link. For example, in a face-to-face mixed-gender
dyadic study, Garcia, Earnshaw, and Quinn (2016) found that men's
reported objectification of their female interaction partner during an
interaction was associated with increased self-objectification reported
by their partner. Gervais, Sáez, Riemer, and Klein (2020) has reviewed
the literature on interpersonal self-objectification and has organized
the the process in a theoretical model called the SIMO, but this model
is only relevant for mixed-gender encounters. Although there is ample
evidence that women objectify other women (Loughnan et al., 2015a; Puvia
\& Vaes, 2013), there is currently no studies investigating the process
of interpersonal objectification same-gender interactions. Does
objectification by a female interaction partner have the same downstream
negative consequences as being objectified by a male interaction
partner? For example, perhaps women could be amplifying each others'
self-objectification as they get ready to go out for the night,
\emph{before} the theoretical encounters with men outlined in Gervais et
al. (2020) occur.

@@

Self-objectification is a multidimensional process that accounts for the
cognitive mechanism that translates experiences of sexualization at the
cultural level (Loughnan et al., 2015a) to psychological (e.g., anxiety,
self-esteem, authenticity, motivational states) and behavioral
(cognitive performance, body monitoring) features of mental health and
well-being at the individual level (Calogero, Tantleff-Dunn, \&
Thompson, 2011; Moradi \& Huang, 2008). Calogero et al. (2011) proposes
that the construct of self-objectification can be conceptualized as a
learned trait. Furthermore, it can also be elicited momentarily, through
the media, for example, with sexualized images in movies and magazines,
which can lead to a state of self-objectification (Bernard, Legrand, \&
Klein, 2018; Calogero et al., 2011; Moradi \& Huang, 2008). Being
objectified by another person and possessing trait-level
self-objectification (TSO) may interact to influence experiences of
feeling like a body, or state self-objectification (SSO) (B. L.
Fredrickson, Roberts, Noll, Quinn, \& Twenge, 1998, Garcia et al.
(2016)). Does this habit self-objectification lead women to also
objectify other women, possibly to compare oneself?

@@

Some studies have found that women objectify other women to a greater
extent than they objectify men (Loughnan et al., 2015a). Women objectify
(dehumanized) sexualized targets presented as images (Puvia \& Vaes,
2013), but we know next to nothing about interpersonal sexual
objectification among women and what the immediate effects might be.
Further, do women objectify non-sexualized women they are interacting
with? What might be different for same-gender interactions? Harsey and
Zurbriggen (2020) found that self-objectification was related to the
objectification of women to a similar degree for men and women
participants, although men scored high than women on a measure of belief
that objectification of women is natural for men.

@@

Self-objectification has been found to occur after a mere relationship
prime among women (Sanchez \& Broccoli, 2008) because in our culture
women need to look attractive to obtain and maintain successful
relationship, thus, the ``college relationships condition may heighten
self-objectification and the evaluation of other women in a sexualized
way. Researchers have also studied interpersonal objectification and
self-objectification in romantic relationships (Strelan \& Pagoudis,
2018). Evidence is mixed, and women can sometimes have positive outcomes
from appearance valuation within the context of an established romantic
relationship and they experience less SSO after appearance comments
within romantic relationships (Meltzer, 2020), less is known about first
dates. But this literature is also only about heterosexual
relationships. A recent study has investigated the effect of
self-objectification on the reduction of relationship building skills in
general (including same-sex friendships) {[}CITE PWQ paper I reviewed,
Yoder editor{]}. Empirical evidence reveals that objectification
manifests through inauthenticity in romantic relationships (Brunell et
al., 2010), adverse attitudes in regard to career aspirations, and a
decrease in concentration and impairment in female cognitive performance
(Kahalon, Shnabel, \& Becker, 2018; D. M. Quinn, Chaudoir, \& Kallen,
2011). Both romantic relationships and relationships in general have
been studies in the context of objectification, but neither of these
literatures have investigated specifically same-gender relationships.

There is evidence that the experience of state self-objectification in
mixed-gender contexts (stranger and romantic) and within the self in any
moment (trying on a swimsuit) has negative consequences, but what about
in the context of interaction with other women? Is the
self-objectification experience in mixed-gender interactions the same as
that experienced in same-gender interactions? The ample research
demonstrating that the male gaze has a particularly detrimental effect
would suggest no (Gay \& Castano, 2010; Gervais, Vescio, \& Allen, 2011;
Saguy, Quinn, F Dovidio, \& Pratto, 2010, Calogero (2004)). Women do
objectify other women (Harsey \& Zurbriggen, 2020; Loughnan et al.,
2015a; Puvia \& Vaes, 2013), but when they do it, does it lead to
self-objectification in the same way men's objectification of women does
(Garcia et al., 2016)? If so, does the self-objectification experienced
in these same-gender interactions have the same negative consequences
for authenticity in that ineraction?

@@

There is some evidence that reduced authenticity is a consequence of
self-objectification in the moment {[}Garcia et al. (2016); again CITE
PWQ paper I reviewed, Yoder editor{]}. This link has been justified by
the literature on stigmatized-stigmatizer interactions, viewing the
experience of being objectified in an interaction as an identity threat
situation. When a woman is objectified by a man, and then experiences
self-objectification, this intergroup encounter might trigger identitiy
threat, but is the same negative consequence present when a woman is
objectified by another woman? woman-woman interpersonal objectification
processes might diverge from this process. One the one hand, she may not
be having the same negative consequences the cascade from situations
that threated group-based identity (Deaux \& Major, 1987; Dovidio, Hebl,
Richeson, \& Shelton, 2006; M. R. Hebl \& Dovidio, 2005), but feeling
like a body, rather than a full human, in an interaction may be enough
to reduce feelings of authenticity and social competence, regardless of
the gender of the objectifyer {[}Tolman, Impett, Tracy, and Michael
(2006); PWQ paper I reviewed, Yoder editor{]}.

@@

Data from two different, but demographically equivalent, samples were
combined to create the final analysis sample (\emph{N =} 64) used in
this study. In the measures section that follows we refer to them as
Sample 1 and Sample 2. Sample 1 (\emph{N =} 24) is from a co-ed liberal
arts college in the northeast US and Sample 2 (\emph{N =} 40) is from a
women's liberal arts college in the northeast US. The description of the
samples section below contains more detail about equivalence analyses to
support the decision to combine these two samples.

@@

\subsubsection{Description of the
Samples}\label{description-of-the-samples}

Thirty-two previously unacquainted self-identifying female-sex dyads (64
total participants) from two liberal arts institutions in the Northeast
of the United States participated in this study. More specifically,
twelve of the pairs, which derived from Sample 1, were students at a
co-ed liberal arts college, while the remaining twenty pairs who came
from Sample 2 attended a women's liberal arts college. Initially, data
was collected from same-sex and mixed-sex dyads that comprised of male
and female gendered individuals. Sample 1 originally consisted of
twenty-two pairs, twelve men and thirty-two women. Twenty-three pairs
made up of forty-three women and one man, as well as two participants
who did not identify with either gender category, formed Sample 2. For
consistency, we limited participant data to same sex female pairs at the
two colleges.

@@

@@

The sample of the current study was comprised of Western women, being
that sexual objectification is most prevalent in this culture (Loughnan
et al., 2015b), and research on objectification conducted outside of
Western or Westernized countries is scarce (Moradi \& Huang, 2008).
Because \enquote{bodies exist within social and cultural contexts, and
hence are also constructed through sociocultural practices and
discourses} (Roberts \& Fredrickson, 1997, p. 174), it is important to
consider how diverse social identities within unique cultural contexts
may inform sexual objectification phenomenon to test the cross-cultural
applicability of theoretical frameworks (Loughnan et al., 2015b).
Further, sexualzing experiences and self-objectification are thought to
begin a very young age, and thus, researchers have only recently begun
to examine such experiences among children (Bury, Tiggemann, \& Slater,
2016; e.g., Holland \& Haslam, 2016; Jongenelis, Byrne, \& Pettigrew,
2014). Considering the fact that the average mean age of the
investigated participants of this current study was 18.85 years,
research among younger and older individuals is needed, especially
because self-objectification may change over time (Roberts \&
Fredrickson, 1997). It may be valuable to question the extent to which
children, adolescents, or emerging adults of different races or
ethnicities are exposed to varied amounts of sexulizing content.

@@

@@

@@

@@

>>>>

Roberts and Fredrickson (1997) Objectification Theory, suggests that
women are objectified in interpersonal encounters, and that the effect
might be strongest from men, but possibly also from women. Psychological
researchers studying the sexual objectification of women, and the
resulting self-objectification that follows, have in recent years
explored the interpersonal process of the objectification -
self-objectification link (CITE INTERPERSONAL PROCESS LINKS LITERATURE).
For example, in a face-to-face mixed-gender dyadic study, Garcia,
Earnshaw, and Quinn (2016) found that men's reported objectification of
their female interaction partner during an interaction was associated
with increased self-objectification reported by their partner. Gervais,
Sáez, Riemer, and Klein (2020) has reviewed the literature on
interpersonal self-objectification and has organized the the process in
a theoretical model called the SIMO, but this model is only relevant for
mixed-gender encounters. Although there is ample evidence that women
objectify other women (Loughnan et al., 2015a; Puvia \& Vaes, 2013),
there is currently no studies investigating the process of interpersonal
objectification same-gender interactions. Does objectification by a
female interaction partner have the same downstream negative
consequences as being objectified by a male interaction partner? For
example, perhaps women could be amplifying each others'
self-objectification as they get ready to go out for the night,
\emph{before} the theoretical encounters with men outlined in Gervais et
al. (2020) occur.

@@

Self-objectification is a multidimensional process that accounts for the
cognitive mechanism that translates experiences of sexualization at the
cultural level (Loughnan et al., 2015a) to psychological (e.g., anxiety,
self-esteem, authenticity, motivational states) and behavioral
(cognitive performance, body monitoring) features of mental health and
well-being at the individual level (Calogero, Tantleff-Dunn, \&
Thompson, 2011; Moradi \& Huang, 2008). Calogero et al. (2011) proposes
that the construct of self-objectification can be conceptualized as a
learned trait. Furthermore, it can also be elicited momentarily, through
the media, for example, with sexualized images in movies and magazines
(Morry \& Staska, 2001), which can lead to a state of
self-objectification (Bernard, Legrand, \& Klein, 2018; Calogero et al.,
2011; Moradi \& Huang, 2008). Being objectified by another person and
possessing trait-level self-objectification (TSO) may interact to
influence experiences of feeling like a body, or state
self-objectification (SSO) (B. L. Fredrickson, Roberts, Noll, Quinn, \&
Twenge, 1998, Garcia et al. (2016)). Does this habit
self-objectification lead women to also objectify other women, possibly
to compare oneself?

@@

Past research has focused on men's objectifying behaviors, with little
attention to the role that women may also play in enacting
objectification of women. Hill and Fischer (2008) assessed women's
experiences of objectification from men separately from women's
experiences of objectification from women and found that women may be
socialized not only to see themselves as objects but perhaps to see
other women as objects as well (THEY DIDN'T DETERMINE THIS IN PERSON).
Some studies have found that women objectify other women to a greater
extent than they objectify men (Loughnan et al., 2015a). Women objectify
(dehumanized) sexualized targets presented as images (Puvia \& Vaes,
2013), but we know next to nothing about interpersonal sexual
objectification among women and what the immediate effects might be.
Further, do women objectify non-sexualized women they are interacting
with? What might be different for same-gender interactions? Harsey and
Zurbriggen (2020) found that self-objectification was related to the
objectification of women to a similar degree for men and women
participants, although men scored high than women on a measure of belief
that objectification of women is natural for men. ASK RANDI-WE KNOW
OBJECTIFICATION HAS MORE ADVERSE CONSEQUENCES FOR WOMEN THAN MEN
(Gervais, Vescio, \& Allen, 2011; Moradi \& Huang, 2008; Saguy, Quinn,
Dovidio, \& Pratto, 2010)\ldots{}MENTION THAT\ldots{}HOWEVER DOES WOMEN
OBJECT WOMEN TO A GREATER EXTENT THAN MEN MATTER MORE?\ldots{}STILL
TRYING TO FIGURE THAT OUT? ADDITIONALLY Strelan and Hargreaves (2005)
FOUND THAT THE MORE WOMEN SELF-OBJECTIFY, THE MORE THEY OBJECTIFY OTHER
WOMEN.

@@

Self-objectification has been found to occur after a mere relationship
prime among women (Sanchez \& Broccoli, 2008) because in our culture
women need to look attractive to obtain and maintain successful
relationship, thus, the ``college relationships condition may heighten
self-objectification and the evaluation of other women in a sexualized
way. Among heterosexual male and female college students, self-
objectification has been found to be positively associated with the
extent to which they objectify their romantic partners (Zurbriggen,
Ramsey, \& Jaworski, 2011). Researchers have also studied interpersonal
objectification and self-objectification in romantic relationships
(Strelan \& Pagoudis, 2018). Evidence is mixed, and women can sometimes
have positive outcomes from appearance valuation within the context of
an established romantic relationship and they experience less SSO after
appearance comments within romantic relationships (Meltzer, 2020), less
is known about first dates. But this literature is also only about
heterosexual relationships. A recent study has investigated the effect
of self-objectification on the reduction of relationship building skills
in general (including same-sex friendships) {[}CITE PWQ paper I
reviewed, Yoder editor{]}. Empirical evidence reveals that
objectification manifests through inauthenticity in romantic
relationships (Brunell et al., 2010), adverse attitudes in regard to
career aspirations, and a decrease in concentration and impairment in
female cognitive performance (Kahalon, Shnabel, \& Becker, 2018; D. M.
Quinn, Chaudoir, \& Kallen, 2011). Both romantic relationships and
relationships in general have been studies in the context of
objectification, but neither of these literatures have investigated
specifically same-gender relationships.

There is evidence that the experience of state self-objectification in
mixed-gender contexts (stranger and romantic) and within the self in any
moment (trying on a swimsuit) has negative consequences, but what about
in the context of interaction with other women? Is the
self-objectification experience in mixed-gender interactions the same as
that experienced in same-gender interactions? The ample research
demonstrating that the male gaze has a particularly detrimental effect
would suggest no (Gay \& Castano, 2010; Gervais, Vescio, \& Allen, 2011;
Saguy, Quinn, F Dovidio, \& Pratto, 2010, Calogero (2004)). Women do
objectify other women (Harsey \& Zurbriggen, 2020; Loughnan et al.,
2015a; Puvia \& Vaes, 2013), but when they do it, does it lead to
self-objectification in the same way men's objectification of women does
(Garcia et al., 2016)? If so, does the self-objectification experienced
in these same-gender interactions have the same negative consequences
for authenticity in that interaction?

@@

There is some evidence that reduced authenticity is a consequence of
self-objectification in the moment {[}Garcia et al. (2016); again CITE
PWQ paper I reviewed, Yoder editor{]}. This link has been justified by
the literature on stigmatized-stigmatizer interactions (CITE LITERATURE
DIDN'T COME ACROSS THIS), viewing the experience of being objectified in
an interaction as an identity threat situation (MAYBE CITE MILES-MCLEAN
PAPER \enquote{STOP LOOKING AT ME!}). When a woman is objectified by a
man, and then experiences self-objectification, this intergroup
encounter might trigger identitiy threat, but is the same negative
consequence present when a woman is objectified by another woman?
woman-woman interpersonal objectification processes might diverge from
this process. One the one hand, she may not be having the same negative
consequences the cascade from situations that threated group-based
identity (Deaux \& Major, 1987; Dovidio, Hebl, Richeson, \& Shelton,
2006; M. R. Hebl \& Dovidio, 2005), but feeling like a body, rather than
a full human, in an interaction may be enough to reduce feelings of
authenticity and social competence, regardless of the gender of the
objectifyer {[}Tolman, Impett, Tracy, and Michael (2006); PWQ paper I
reviewed, Yoder editor{]}. I DONT KNOW WHERE THIS PAPER IS? ASK RANDI
ABOUT IDENTITY\_THREAT PAPERS

@@

Data from two different, but demographically equivalent, samples were
combined to create the final analysis sample (\emph{N =} 64) used in
this study. In the measures section that follows we refer to them as
Sample 1 and Sample 2. Thirty-two previously unacquainted
self-identifying female-sex dyads (64 total participants) from two
liberal arts institutions in the Northeast of the United States
participated in this study. Sample 1 (\emph{N =} 24) is from a co-ed
liberal arts college in the northeast US and Sample 2 (\emph{N =} 40) is
from a women's liberal arts college in the northeast US. More
specifically, twelve of the pairs, which derived from Sample 1, were
students at a co-ed liberal arts college, while the remaining twenty
pairs who came from Sample 2 attended a women's liberal arts college.
Initially, data was collected from same-sex and mixed-sex dyads that
comprised of male and female gendered individuals. Sample 1 originally
consisted of twenty-two pairs, twelve men and thirty-two women.
Twenty-three pairs made up of forty-three women and one man, as well as
two participants who did not identify with either gender category,
formed Sample 2. For consistency, we limited participant data to same
sex female pairs at the two colleges.

@@

@@

The current study tested whether the model of interpersonal
objectification and state self-objectification (SSO) used in Garcia et
al. (2016) replicates in a same-gender dyadic sample of women only.
Although past reserach has found that women do objectify other women
(Harsey \& Zurbriggen, 2020; Loughnan et al., 2015a; Puvia \& Vaes,
2013), this is the first study to test if this other-objectification by
women is related to self-objectification in women. As hypothesized, the
current study did find a significant relationship bewteen a woman's
partner's report of having objectified her and her own post-interaction
feelings of self-objectification.

Where the results of the current study diverge most notably from the
results of studies testing interpersonal objectification among
mixed-gender dyads is the lack of evidence for relationships between SSO
and felt inauthenticity (Garcia et al., 2016). In addition, we found to
evidence of \emph{direct} or \emph{indirect} relationships between SSO
and cognitive functioning, relationship agency, and career aspirations.
Nor was there evidence of \emph{direct} or \emph{indirect} relationships
between other-objectification and any of these outcomes. This is
somewhat suprising given the extant research on interpersonal
other-ojectification and cognitive functioning {[}CITE that logel
study{]}, and the plentiful evidence linking SSO and cognitive
functioning {[}B. L. Fredrickson et al. (1998); CITE; CITE; CITE{]}.
This lack of evidence could potentially signal diverging processs
between women's experiences with interpersonal objectification from men
and interpersonal objectification from women. There is quite a bit of
evidence suggesting that the male gaze is particularly detrimental
{[}CITE; CITE; CITE{]}, and perhaps the self-objectification experienced
within an interaction with a women is not as detrimental as the
self-objectificatoin experienced within an interaction with a man. This
lack of evidence could also be due to the relatively small sample size.

What we did find, however,

@@

The sample of the current study was comprised of Western women, being
that sexual objectification is most prevalent in this culture (Loughnan
et al., 2015b), and research on objectification conducted outside of
Western or Westernized countries is scarce (Moradi \& Huang, 2008),
although more current work examines objectification from a
cross-cultural framework (Loughnan et al., 2015b, Wollast et al.
(2020)). Because \enquote{bodies exist within social and cultural
contexts, and hence are also constructed through sociocultural practices
and discourses} (Roberts \& Fredrickson, 1997, p. 174), it is important
to consider how diverse social identities within unique cultural
contexts may inform sexual objectification phenomenon to test the
cross-cultural applicability of theoretical frameworks (Loughnan et al.,
2015b). Further, sexualizing experiences and self-objectification are
thought to begin a very young age, and thus, researchers have only
recently begun to examine such experiences among children (Bury,
Tiggemann, \& Slater, 2016; e.g., Holland \& Haslam, 2016; Jongenelis,
Byrne, \& Pettigrew, 2014). Considering the fact that the average mean
age of the investigated participants of this current study was 18.85
years, research among younger and older individuals is needed,
especially because self-objectification may change over time (Roberts \&
Fredrickson, 1997). It may be valuable to question the extent to which
children, adolescents, or emerging adults of different races or
ethnicities are exposed to varied amounts of sexualizing content.

@@

\hypertarget{ref-hill2008examining}{}
Hill, M. S., \& Fischer, A. R. (2008). Examining objectification theory:
Lesbian and heterosexual women's experiences with sexual-and
self-objectification. \emph{The Counseling Psychologist}, \emph{36}(5),
745--776.

@@

\hypertarget{ref-morry2001magazine}{}
Morry, M. M., \& Staska, S. L. (2001). Magazine exposure:
Internalization, self-objectification, eating attitudes, and body
satisfaction in male and female university students. \emph{Canadian
Journal of Behavioural Science/Revue Canadienne Des Sciences Du
Comportement}, \emph{33}(4), 269.

@@

\hypertarget{ref-wollast2020cultural}{}
Wollast, R., Riemer, A. R., Gervais, S. J., Grigoryan, L., Bernard, P.,
\& Klein, O. (2020). How cultural orientation and self-compassion shape
objectified body consciousness for women from america, belgium, russia,
and thailand. \emph{Self and Identity}, 1--21.

@@

\hypertarget{ref-zurbriggen2011self}{}
Zurbriggen, E. L., Ramsey, L. R., \& Jaworski, B. K. (2011). Self-and
partner-objectification in romantic relationships: Associations with
media consumption and relationship satisfaction. \emph{Sex Roles},
\emph{64}(7-8), 449--462.
